Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-07 paper/paper/SPIE/article.tex
[WIP]

% \author[a]{First Author}
% \author[a]{Second Author}
% \author[b]{Third Author}
% \author[a,b,*]{Fourth Author}
% \affil[a]{University Name, Faculty Group, Department, Street Address, City, Country, Postal Code}
% \affil[b]{Company Name, Street Address, City, Country, Postal Code}

@@

At its most rudimentary, RMT simply describes the statistically-expected
behaviour of the eigenvalues of some classes—or \textit{ensembles}—of random
matrices under certain assumptions or conditions. A random matrix is,
generally, a matrix in which the entries are drawn from independently
and identically distributed (\textit{iid}) distributions. For example, one of the simplest
ensembles [cite handbook] is the Gaussian Unitary Ensemble (GUE), wherein [desc
of ensemble properties] and which appears to be useful in describing
[real-world examples].

This means that if we examine a system and find that the distribution of
eigenvalues of that system resemble some RMT predicted spectrum, we potentially
gain insight into the nature of that system. If, in addition, we find a large
deviance from the spectrum predicted by RMT, we find contrapositive evidence
that the system is \textit{not} like those described by RMT. Roughly, complex
systems well-described by RMT are highly random and independent, whereas
systems that are not so well-described are either highly dependent, non-random,
or both.

In this sense, RMT might have real \textit{explanatory} and
\textit{statistical} value: the more we see that a system resembles an RMT
system, the less we can distinguish it from one of random, non-interacting
components, and the less the system resembles an RMT system, the more
fundamentally complex and interactive the interactions.

For a system as complex as the human brain, RMT is thus tantalizing. If some
matrix representation of brain activity can be described, even partially, by
aspects of RMT theory, this would mean  description of this matrix either aligned nicely with
RMT theory, or otherwise was broadly predictively useful.

[Transition to fMRI now...]

In functional magnetic resonance imaging (fMRI), changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity. It is common to investigate statistical relationships between the
BOLD signals through \textbf{functional connectivity} analyses, where
correlations between collections of these signals are examined to infer
connections between different voxels or regions of interest (ROIs) within the
brain.

Whether in the presence of experimental stimuli, or the relative absence, as in
resting state fMRI (rs-fMRI), complex patterns in functional connectivity
networks are ubiquitous
\cite{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}.
This complexity suggests fMRI is a candidate to be studied using \textbf{Random
Matrix Theory} (RMT), a set of mathematical tools originally developed some 50
years ago to solve complex problems in nuclear physics
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}

In various physical systems where nuclei can have different possible states,
precisely modeling the entirety of the interactions between nuclei is often
intractable. For such large, complex systems, the eigenvalues (or spectra)
efficiently summarize the totality of the interactions between components of
the system, and RMT describes the expected behaviour of such eigenvalues. By
analysing the spectrum statistically, complex systems can be investigated by
comparison to the universal properties predicted by RMT.

@@

Recently, RMT has been applied to study brain functioning. The earliest study
demonstrated that spectra of the correlations between electroencephalographic
(EEG) signals closely resemble those of the Gaussian Orthogonal Ensemble (GOE)
\cite{sebaRandomMatrixAnalysis2003}. Somewhat more recently, RMT was used to
evaluate the quality of whole brain features extracted from fMRI data
\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}. RMT has
also been used in diffusion MRI to aid in the selection of the number of
components to employ in principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

@@

@@

The level number variance, \(\Sigma^2(L)\), or level variance, for short, is
closely related to the spectral rigidity \cite{mehtaRandomMatrices2004}, and is
calculated as:
\begin{equation}
\label{eq:levlvar}
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\end{equation}
where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\),
and where \(c\), \(L\), and \(\langle \cdot \rangle_c\) are as above
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as an \textit{irregular} timeseries, the level number varaince is simply the
variance of the number of samples of all intervals of length \(L\) over the
series.

@@

\subsection{Tables}
Tables are numbered in the order in which they are referenced. They should appear in the document in numerical order and as close as possible to their first reference in the text. It is preferable to have tables appear at the top or bottom of the page, if possible. Table captions are handled identically to those for figures, except that they appear above the table. See Table~\ref{tab:fonts} for an example.

\subsection{Multimedia}
Acceptable file formats, including MOV (.mov), MPEG (.mpg), and MP4 (.mp4), are playable using standard media players, such as VLC or Windows Media Player. The recommended maximum size for each multimedia file is 10-12 MB. Authors must insert a representative still image from the video file in the manuscript as a figure. The caption label will be linked by the publisher to the actual video file. The video may also be mentioned in an existing figure caption. Multimedia files are treated in the same manner as figures and they will be numbered sequentially with normal figures.  The video number, file type, and file size should be included in parentheses at the end of the figure caption. See Figure \ref{vid:satellite} for an example.

\begin{video}
\begin{center}
{\includegraphics[height=5cm]{satellite.eps}}
\\
\end{center}
\caption{\label{vid:satellite}This satellite is a still image from Video 1 (Video 1, MPEG, 2.5 MB).}
\end{video}

@@

\section{Miscellaneous Formatting Details}
\label{sect:misc}
At times it may be desired, for formatting reasons, to break a line without starting a new paragraph. In a LaTeX source file, a linebreak is created with \verb|\\|.

\subsection{Formatting Equations}
Equations may appear inline with the text, if they are simple, short, and not of major importance; for example, $\beta = b/r$.  Important equations appear on their own line.  Such equations are centered.  For example, ``The expression for the field of view is
\begin{equation}
\label{eq:fov}
2 a = \frac{(b + 1)}{3c} \, ,
\end{equation}
where $a$ is the ...''  Principal equations are numbered, with the equation number placed within parentheses and right justified.

Equations are considered to be part of a sentence and should be punctuated accordingly. In the above example, a comma appears after the equation because the next line is a subordinate clause. If the equation ends the sentence, a period should follow the equation. The line following an equation should not be indented unless it is meant to start a new paragraph. Indentation after an equation is avoided in LaTeX by not leaving a blank line between the equation and the subsequent text.

References to equations include the equation number in parentheses, for example, ``Equation~(\ref{eq:fov}) shows ...'' or ``Combining Eqs.~(2) and (3), we obtain...'' Note that the word ``Equation'' is spelled out if it begins a sentence, but is abbreviated as ``Eq.'' otherwise. Using a tilde in the LaTeX source file between two characters avoids unwanted line breaks, for example between ``Eq.'' and the following equation number..

\subsection{Formatting Theorems}

To include theorems in a formal way, the theorem identification should appear in a 10-point, bold font, left justified, and followed by a period.  The text of the theorem continues on the same line in normal, 10-pt. font, achieved in LaTeX using \verb|\footnotesize|.  For example,

\vspace{2ex}\noindent{\footnotesize\textbf{Theorem 1.} For any unbiased estimator...}

>>>>

@@

At its most rudimentary, RMT describes the expected behaviour of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions.

RMT likely has the most real-world explanatory potential when dealing with a
complex system of many (hundreds or more\cite{mehtaRandomMatrices2004})
interacting components. If such a system has a matrix representation, and the
eigenvalues of this representation have a distribution similar to one predicted
by RMT, it suggests the system is either highly random or chaotic. By contrast,
if the observed spectra deviate significantly from RMT-predicted spectra, this
suggests otherwise. A number of studies have used RMT to make such
interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

RMT may thus have real potential in understanding the human brain. For an
imaging modality like functional magnetic resonance imaging (fMRI), where
changes in the blood-oxygenation-level-dependent (BOLD) signals are related to
neural activity, RMT may be an ideal starting point. Each voxel or collection
of voxels—i.e. Region of Interest (ROI\footnote{We will use ROI to refer to
\textit{any} collection of voxels of any size. E.g. a single voxel is still an
ROI.}); any collection )—can be considered an interacting component of the system.

Likewise, in \textbf{functional connectivity} analyses, statistical
relationships of the BOLD ROI time-courses are investigated in the hope of
gaining insights into brain functioniong (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

The earliest study taking this approach
demonstrated that spectra of the correlations between electroencephalographic
(EEG) signals closely resemble those of the Gaussian Orthogonal Ensemble (GOE)
\cite{sebaRandomMatrixAnalysis2003}. In fMRI, RMT has been used to
evaluate the quality of whole brain features extracted from fMRI data
\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}, and in
diffusion MRI to aid in the selection of the number of
components to employ in principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

@@

\subsubsection{Raw Eigenvalues}
\label{sec:raw-eigs}

The time and space complexities of computing the full \(N \times N\) correlation
matrix of \(\mathbf{M}\), and the subsequent eigenvalues, are too large to be
tractable. However, since transposition does not change the eigenvalues, we
can use the transpose to efficiently compute these eigenvalues (see
Appendix \ref{sec:transpose}).

@@

The level number variance, \(\Sigma^2(L)\), or level variance, for short, is
closely related to the spectral rigidity \cite{mehtaRandomMatrices2004}, and is
calculated as:
\begin{equation}
\label{eq:levlvar}
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\end{equation}
where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\),
and where \(c\), \(L\), and \(\langle \cdot \rangle_c\) are as above
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as an \textit{irregular} timeseries, the level number variance is the
variation of the number of samples of all intervals of length \(L\) over the
series.

@@

% \subsection{Tables}
% Tables are numbered in the order in which they are referenced. They should appear in the document in numerical order and as close as possible to their first reference in the text. It is preferable to have tables appear at the top or bottom of the page, if possible. Table captions are handled identically to those for figures, except that they appear above the table. See Table~\ref{tab:fonts} for an example.

% \subsection{Multimedia}
% Acceptable file formats, including MOV (.mov), MPEG (.mpg), and MP4 (.mp4), are playable using standard media players, such as VLC or Windows Media Player. The recommended maximum size for each multimedia file is 10-12 MB. Authors must insert a representative still image from the video file in the manuscript as a figure. The caption label will be linked by the publisher to the actual video file. The video may also be mentioned in an existing figure caption. Multimedia files are treated in the same manner as figures and they will be numbered sequentially with normal figures.  The video number, file type, and file size should be included in parentheses at the end of the figure caption. See Figure \ref{vid:satellite} for an example.

% \begin{video}
% \begin{center}
% {\includegraphics[height=5cm]{satellite.eps}}
% \\
% \end{center}
% \caption{\label{vid:satellite}This satellite is a still image from Video 1 (Video 1, MPEG, 2.5 MB).}
% \end{video}

@@

\section{Correlation Eigenvalues via Transposition}
\label{sec:transpose}

Let \(\mathbf{X}\) is a real \(n \times p\) matrix with \(n > p\). Let
\[
\mathbf{Z} = \text{norm}(\mathbf{X}) =
\left(
\mathbf{X}_1 - \bar{\mathbf{X}}_1 | \dots | \mathbf{X}_p - \bar{\mathbf{X}}_p
\right)
\]
where \(\mathbf{X}_i\) denotes column \(i\) of \(\mathbf{X}\). Denote the (unordered) set
of eigenvalues of \(\mathbf{X}\) as \(\text{eigs}(\mathbf{X})\), and let \(r = (p - 1)^{-1}\).
Denote the covariance matrix of \(\mathbf{X}\) as \(\text{cov}(\mathbf{X})\).  Then:

\begin{align*}
\text{eigs}\left( \text{cov}(\mathbf{X})  \right)
&= \text{eigs}(r \cdot \mathbf{Z}\mathbf{Z}^{\top}) \\
&= r \cdot \text{eigs}(\mathbf{Z}\mathbf{Z}^{\top}) \\
&= r \cdot \text{eigs}((\mathbf{Z}\mathbf{Z}^{\top})^{\top}) \\
&= r \cdot \text{eigs}(\mathbf{Z}^{\top}\mathbf{Z}) \\
\end{align*}

Denote the correlation matrix of \(\mathbf{X}\) as \(\text{corr}(\mathbf{X})\), and let
\[
\mathbf{Y} = \text{stand}(\mathbf{X}) =
\left(
(\mathbf{X}_1 - \bar{\mathbf{X}}_1) / \sigma_1 | \dots | (\mathbf{X}_p - \bar{\mathbf{X}}_p) / \sigma_p
\right)
\]

Then

\[
\text{eigs}\left( \text{corr}(\mathbf{X})  \right)
= \text{eigs}\left( \text{cov}(\mathbf{Y})  \right)
= r \cdot \text{eigs}(\mathbf{Y}^{\top}\mathbf{Y})
\]

% \section{Miscellaneous Formatting Details}
% \label{sect:misc}
% At times it may be desired, for formatting reasons, to break a line without starting a new paragraph. In a LaTeX source file, a linebreak is created with \verb|\\|.

% \subsection{Formatting Equations}
% Equations may appear inline with the text, if they are simple, short, and not of major importance; for example, $\beta = b/r$.  Important equations appear on their own line.  Such equations are centered.  For example, ``The expression for the field of view is
% \begin{equation}
% \label{eq:fov}
% 2 a = \frac{(b + 1)}{3c} \, ,
% \end{equation}
% where $a$ is the ...''  Principal equations are numbered, with the equation number placed within parentheses and right justified.

% Equations are considered to be part of a sentence and should be punctuated accordingly. In the above example, a comma appears after the equation because the next line is a subordinate clause. If the equation ends the sentence, a period should follow the equation. The line following an equation should not be indented unless it is meant to start a new paragraph. Indentation after an equation is avoided in LaTeX by not leaving a blank line between the equation and the subsequent text.

% References to equations include the equation number in parentheses, for example, ``Equation~(\ref{eq:fov}) shows ...'' or ``Combining Eqs.~(2) and (3), we obtain...'' Note that the word ``Equation'' is spelled out if it begins a sentence, but is abbreviated as ``Eq.'' otherwise. Using a tilde in the LaTeX source file between two characters avoids unwanted line breaks, for example between ``Eq.'' and the following equation number..

% \subsection{Formatting Theorems}

% To include theorems in a formal way, the theorem identification should appear in a 10-point, bold font, left justified, and followed by a period.  The text of the theorem continues on the same line in normal, 10-pt. font, achieved in LaTeX using \verb|\footnotesize|.  For example,

% \vspace{2ex}\noindent{\footnotesize\textbf{Theorem 1.} For any unbiased estimator...}
